---
title: "Collection timepoint distribution"
author: "Christian Santos-Medellin"
---

The only point of this notebook is to generate a plot that displays all the samples that were collected for this study. 
This could be used as a supplementary figure that serves as reference to contextualize where the data for some analyses is coming from. 

Load libraries
```{r}
source("../../General/general_functions.R")
source("../../General/plotting_parameters.R")
library(lubridate)
library(cowplot)
library(tidyverse)
```

Load data
```{r}
climate <- read.table("../Data/climate.tsv", header = T, sep = "\t") %>% 
  mutate(Date = mdy(Date))

soil.map <- readRDS("../Data/soil_map.RDS")

map <- readRDS("../Data/dna_map.RDS") %>% 
  mutate(SampleID = DNAID) %>% 
  left_join(select(soil.map, SoilID, GSM:Na_Sat), by = "SoilID") 
```

Set some parameters for plotting
```{r}
# The min and max dates define the period that will be displayed for the plot
min.date <- "2020-10-01"
max.date <- "2022-03-01"

# this data frame will be used to highlight each of the 3 years encompassed by the sampling campaign
period.df <- tribble(
  ~Period, ~PeriodLong, ~Start,  ~End, ~Mid,
  "P1", "2020", min.date, "2020-12-31", "2020-11-15",
  "P2", "2021", "2021-01-01", "2021-12-31", "2021-07-01",
  "P3", "2022", "2022-01-01", max.date, "2022-02-01"
) %>% 
  mutate(StartDate = ymd(Start),
         EndDate = ymd(End),
         MidDate = ymd(Mid))

# this vector contains the abbreviated months to display in the x axis 
# NOTE that if you change the min and max dates, this vector will have to be updated
year_labels <- c('O','N','D','J','F','M','A','M','J','J','A','S','O','N','D','J','F',"M")
```

Plot the precipitation patterns observed during the sampling period
For this plot, I am displaying the actual max values observed in the study (which shrinks most of the trends given the record ppt we had at the end of 2021).
An alternative version of this precipitation data with trimmed values is shown in the mound_vs_swale.Rmmd notebook 
```{r}
ppt.p <- climate %>% 
  filter(Date > as.Date(min.date)) %>%
  filter(Date < as.Date(max.date)) %>%
  ggplot() +
  geom_line(aes(Date, TotalPrecipitation)) +
  geom_rect(data = period.df, aes(xmin = StartDate, xmax = EndDate + 1, 
                                  ymin = -2, ymax = -7, 
                                  fill = Period)) + 
  geom_text(data = period.df, aes(x = MidDate, label = PeriodLong), y = -4.5, hjust = 0.5, vjust = 0.5) + 
  geom_point(data = soil.map %>% filter(Study == "Temporal") %>% group_by(CollectionDate) %>% count(), aes(CollectionDate, - 8, color = as.factor(CollectionDate))) +
  scale_color_manual(values = timepoint.pal) +
  scale_fill_manual(values = c("gray25", "gray75", "gray25")) +
  scale_x_date("Date",breaks = c(seq(from=as.Date(min.date),to=as.Date(max.date),by="month")),
               labels = year_labels)  +
  ylab("Daily rainfall\n(mm)") +
  basic.theme() +
  theme(legend.position = "none")

ppt.p  
```

Plot a heatmap that shows how many replicates were generated for each sample type in each time point. 
The plot also includes the color codes used for each time point
```{r}
# Create a data frame summarizing each collection time point and the associated date
# The extra variables are just there so we can rbind them with the other data frames. 
time.tmp <- soil.map %>% 
  group_by(CollectionDate, Timepoint) %>% 
  filter(Study == "Temporal") %>% 
  count() %>% 
  ungroup() %>% 
  arrange(CollectionDate) %>% 
  mutate(tmp = 1:n()) %>% 
  mutate(Feature = "tmp",
         Title = "tmp",
         Profile2 = "tmp",
         Replicates = "tmp") %>% 
  select(CollectionDate, Timepoint, Feature, Title, Profile2, Replicates)

# Create a data frame summarizing the number of sequenced profiles generated for each time point
seq.tmp <- map %>% 
  mutate(AmpliconSequenced = ifelse(AmpliconSequenced, "Yes", "No"),
         WGSSequenced = ifelse(WGSSequenced, "Yes", "No")) %>% 
  gather(key = "Profile", value = "Value", WGSSequenced, AmpliconSequenced) %>% 
  group_by(Position, Feature, CollectionDate, Profile, Fraction, Value, DNase, SoilStatus, Study) %>% 
  count() %>% 
  ungroup() %>% 
  spread(key = Value, value = n) %>% 
  mutate(No = ifelse(is.na(No), 0, No),
         Yes = ifelse(is.na(Yes), 0, Yes)) %>% 
  filter(Study == "Temporal") %>% 
  filter(!(Profile == "WGSSequenced" & Fraction == "Total")) %>% 
  filter(!(Profile == "AmpliconSequenced" & DNase)) %>% 
  mutate(DNase2 = ifelse(DNase, "+DNase", "Untreated")) %>% 
  mutate(Fraction2 = ifelse(Fraction == "Viral", "0.22 µ DNA", "Total DNA")) %>% 
  mutate(SoilStatus2 = ifelse(SoilStatus == "Fresh", "Fresh soil", "Frozen soil")) %>% 
  mutate(Profile2 = ifelse(Profile == "AmpliconSequenced", "16S rRNA gene\nprofiles", "Virome\nprofiles")) %>% 
  mutate(Title = paste0(SoilStatus2, "\n", Fraction2, " / ", DNase2)) %>% 
  mutate(Replicates = as.factor(Yes)) %>% 
  mutate(Timepoint = NA) %>% 
  select(CollectionDate, Timepoint, Feature, Title, Profile2, Replicates)

# Create a data frame summarizing the number of edaphic profiles generated for each time point
nut.tmp <- soil.map %>% 
  mutate(AbioticProfiled = !is.na(WardBatch)) %>% 
  group_by(Position, Feature, CollectionDate, AbioticProfiled, Study) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(Replicates = ifelse(AbioticProfiled, n, 0) %>% as.factor()) %>% 
  filter(Study == "Temporal") %>% 
  mutate(Profile2 = "Soil nutrient\nprofiles") %>% 
  mutate(Title = "Refrigerated soil") %>% 
  mutate(Timepoint = NA) %>% 
  select(CollectionDate, Timepoint, Feature, Title, Profile2, Replicates)

# Bind data frames and plot
samples.p <- rbind(time.tmp, seq.tmp, nut.tmp) %>% 
  mutate(Title = fct_relevel(Title, "tmp", "Fresh soil\nTotal DNA / Untreated")) %>% 
  mutate(Profile2 = fct_relevel(Profile2, "tmp", "Virome\nprofiles")) %>% 
  ggplot(aes(as.factor(CollectionDate), Feature)) +
  geom_tile(data = . %>% filter(Title == "tmp"), aes(fill = Timepoint), color = "black", size = 0.5) +
  geom_text(data = . %>% filter(Title == "tmp"), aes(label = Timepoint), size = 2) + 
  scale_fill_viridis_d(option = "C") +
  geom_tile(data = . %>%  filter(Title != "tmp" & Replicates == "0"), color = "black", fill = rep.pal[1], size = 0.5) +
  geom_tile(data = . %>%  filter(Title != "tmp" & Replicates == "1"), color = "black", fill = rep.pal[2], size = 0.5) +
  geom_tile(data = . %>%  filter(Title != "tmp" & Replicates == "2"), color = "black", fill = rep.pal[3], size = 0.5) +
  geom_tile(data = . %>%  filter(Title != "tmp" & Replicates == "3"), color = "black", fill = rep.pal[4], size = 0.5) +
  geom_tile(data = . %>%  filter(Title != "tmp" & Replicates == "4"), color = "black", fill = rep.pal[5], size = 0.5) +
  scale_color_viridis_d(option = "C") +
  scale_x_discrete(position = "top") +
  facet_grid(Profile2 + Title ~ ., scale = "free", space = "free") +
  basic.theme() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.title.y = element_blank(),
        strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0),
        legend.position = "none")

samples.p
```

Create a dummy plot that displays the legend for the replicate colors in the previous plot
This is needed because I didn't use replicate fill color as an aesthetic since I used it for the time points.
```{r}
legend.p <- seq.tmp %>% 
  ggplot(aes(as.factor(CollectionDate), Feature, fill = Replicates)) +
  geom_tile(aes(fill = Replicates), color = "black", size = 0.5) +
  scale_fill_manual(values = rep.pal) +
  scale_x_discrete(position = "top") +
  facet_grid(Profile2 + Title ~ ., scale = "free", space = "free") +
  basic.theme() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        axis.title.y = element_blank(),
        strip.text = element_text(size = 8),
        strip.text.y = element_text(angle = 0),
        legend.position = "bottom")

legend.p
```

Put it all together
800:600
```{r}
plot_grid(ppt.p, samples.p, ncol = 1, rel_heights = c(1,2), align = "v", axis = "lr")

plot_grid(get_legend(legend.p))
```

